Guilt, shame, and anti-war action in an authoritarian country at war

### 2024-06-10 manuscript.Rmd
Back to tinytex
Implemented most revisions

---
title: "Guilt, shame, and anti-war action in an authoritarian country at war"
output:
  bookdown::pdf_document2:
    latex_engine: lualatex
    keep_tex: yes
    toc: no
  word_document:
    toc: no
  bookdown::word_document2:
    toc: no
  html_document:
    df_print: paged
always_allow_html: yes
test: "`r OS <- Sys.info()['sysname']`
`r if (OS=='Linux'){fontname <- 'Nimbus Roman'}`
`r if (OS=='Darwin'){fontname <- 'Times New Roman'}`
`r if (OS=='Windows'){fontname <- 'Times New Roman'}`"
header-includes:
- \usepackage{indentfirst}
- \usepackage{sectsty}
- \allsectionsfont{\centering}
- \usepackage{fontspec}
- \usepackage{polyglossia}
- \setdefaultlanguage{english}
- \setotherlanguages{russian}
- \setmainfont{`r fontname`}
- \newfontfamily{\cyrillicfonttt}{`r fontname`}
- \usepackage{float}
- \usepackage{float} \floatplacement{figure}{H}
- \newcommand{\beginsupplement}{\setcounter{table}{0}
- \renewcommand{\thetable}{S\arabic{table}}
- \setcounter{figure}{0}
- \renewcommand{\thefigure}{S\arabic{figure}}}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
indent: yes
bibliography: references.bib
csl: elife.csl
---

\fancyhead[LH]{GUILT, SHAME, AND ANTI-WAR ACTION}

@@

$^{5}$ Department of Psychological and Brain Sciences, University of Massachusetts Amherst
\vspace{50mm}

@@

@@

Studies of group-based guilt and shame show that these emotions can facilitate intergroup reconciliation. However, most of these studies are conducted in democratic countries and usually on past, not present events. We set out to investigate the role of group-based guilt and shame in collective action in an authoritarian country at war. We asked more than 1000 Russians living in Russia, a sample representative of the country’s population by gender and age,  about their experiences of group-based guilt and shame in relation to Russia’s invasion of Ukraine and their past and future anti-war political actions. The study was guided by three research questions: (1) Are feelings of group-based guilt and shame conditional upon political beliefs that people can influence the actions of their governments? (2) Are guilt and shame predictive of anti-war action in an authoritarian state? (3) Are guilt and shame better predictors of anti-war action than other emotions or attitudes? Democratic values, not political efficacy, was the most robust predictor of group-based guilt and shame. Only moral shame, but not guilt or image shame predicted past and future anti-war action. Whereas attitude towards the war and moral shame predicted whether or not participants intend to engage in political action, negative emotions predicted the strength of these intentions. We highlight the gaps in the study of collective action and the need for more evidence from non-democratic contexts.

\vspace{5mm}
*Keywords*: group-based guilt, moral shame, image shame, collective action, Russia-Ukraine war, authoritarianism

@@

packagelist <- c('readr','psych','GPArotation','lavaan','pscl','ggplot2','sjPlot', 'car','dplyr','MASS','writexl','readxl','kableExtra','rstatix','tinytex')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

@@

On February 24, 2022, the country with the biggest nuclear warhead inventory in the world, Russia, invaded the neighboring Ukraine. Although the Russia-Ukraine war started in 2014 with the occupation of Crimea, Russian officials consistently denied their involvement in the region. Eight years later, to the surprise of most analysts and pundits, Russia openly invaded Ukraine on four fronts: North (Kyiv), North-East (Kharkiv), South-East (Donetsk and Luhansk), and South (Crimea). The official ideological narrative supporting the invasion was provided in a televised speech by Vladimir Putin preceding the invasion by mere minutes. In this speech, Putin falsely claimed that Ukraine is governed by a group of neo-Nazis who hold the population hostage and persecute the Russian speakers in the country. The announced goals of this “special military operation” (the use of the word “war” soon after will be criminalized in Russia) were the “denazification” and “demilitarization” of Ukraine.

In the first days following the invasion, thousands of Russians take  to the streets to protest. However, nearly 15.000 people are detained in the first weeks following the invasion [@OVDInfo2022] and the street protests die down quickly. New laws are introduced regularly to further supress protest, among them the law about the “discreditation of the Russian armed forces”, which essentially prohibits any anti-war statements [@Eckel2022]. After the introduction of this law, the few remaining independent media outlets either stop covering the war or leave the country [@MoscowTimes2022].

Against the backdrop of the reports of war crimes committed by the Russian armed forces in Ukraine and the repression of all dissent in Russia, the discussions among the well-educated Russians, many of whom left the country, often revolve around the topics of collective responsibility, guilt, and shame [e.g., @Grigoryan2022; @Treus2022]. As these discussions take shape, it is becoming clear that the majority view, even among the Russian “intelligentsia”, is that feelings of guilt and shame are inappropriate or even harmful [@Kynev2022]. For example, in a series of interviews with prominent Russian intellectuals and artists, Ekaterina Gordeeva, a journalist with more than 1.5 million subscribers on YouTube, asks each of her guests whether they feel guilt or shame for what is happening in the country; with very few exceptions, the answer is “no” [@Gordeeva2022]. The push-back against the idea of feeling guilt or shame for what your country is doing is so strong that soon a new hashtag #IamNotAshamed (#МнеНеСтыдно) becomes popular.

Research on group-based guilt and shame and collective action has been predominantly carried out in Western democracies. Although there is strong evidence that experiencing these emotions can motivate people to engage in collective action and support intergroup reconciliation efforts [@Hakim2021; @Lickel2011], research on preconditions for experiencing these emotions and on the power of these emotions to motivate collective action in authoritarian regimes is lacking. Is the sense that you have some control over the political events in your country a necessary precondition for feeling group-based guilt and shame? Are these feelings relevant in authoritarian regimes? Can they motivate behavior in such contexts? We set out to address these questions in Russia, an authoritarian country at war.

@@

This prediction that guilt would motivate reconciliatory behavior, whereas shame would motivate avoidance is not well supported by the empirical evidence available so far. A recent meta-analysis of group-based emotions ([@Hakim2021], 101 effect sizes, 58 samples, N = 10,305) found that guilt, shame, and anger are equally predictive of support for reparations, with a strong average effect size of *r* = .44 and no significant differences between emotions. Furthermore, the distinction between moral shame and image shame [@Allpress2010; @Rees2013] helps explain which elements of shame are predictive of reconciliatory behaviors and which are predictive of avoidant behaviors. This distinction is made based on different appraisals that the feeling of shame can be accompanied with: we feel moral shame when the group members’ behavior violates important moral standards or values of the group, and we feel image shame when the group members’ behavior tarnishes the ingroup’s reputation [@Rees2013]. Studies that make the distinction between moral shame and image shame find that moral shame is predictive of reconciliatory behaviors on par or more strongly than guilt, whereas image shame is predictive of distancing and avoidance  [@Allpress2010; @Allpress2014; @Grigoryan2017; @Nooitgedagt2021; @Rees2013]. We therefore expect that moral shame and guilt, but not image shame, will predict a stronger intention to act against the war in Russia (H1).

## Acknowledgment of responsibility as a precondition for experiencing guilt and shame {.unnumbered}

Although rarely studied, the acknowledgment of collective responsibility for the ingroup’s wrongdoings is a necessary precondition for experiencing group-based guilt and shame. In a series of experiments, @Cehajic-Clancy2011 showed that guilt mediates the effect of an intervention designed to increase acknowledgment of responsibility for the ingroup’s wrongdoing and support for reparations. Consistent with this finding, a recent set of studies showed that beliefs of group malleability shape experiences of guilt and shame [@Weiss-Klayman2020]: participants who believed that groups can change and those who thought that others believe that groups can change (meta-beliefs) showed higher levels of guilt and shame.

@@

@@

The integrative social identity model of collective action (SIMCA) – perhaps the most comprehensive and widely used model of collective action in psychology – identifies three key components of collective action: perceived injustice, identity, and efficacy [@Zomeren2008]. In a meta-analysis of more than 180 independent samples, @Zomeren2008 find that each of these three components uniquely predict collective action, with similar effect sizes. A recent extension adds moral beliefs as a key component of the model [@Zomeren2018]. However, most studies of collective action have been conducted in democratic countries, where the cost of political participation is not that high. Moreover, most of these studies focus on political movements related to minority rights and social justice and the generalizability of these findings to political movements aimed at regime change in authoritarian states has not been well studied.

In a series of studies in five non-democratic states (Egypt and Turkey in 2013, Russia, Ukraine, and Hong Kong in 2014), Ayanian and colleagues [@Ayanian2016; @Ayanian2021] show how perceptions of risk in repressive contexts can have counter-intuitive effects on willingness to participate in collective action. While perceived risk can quell collective action through increased fear, it can also spur resistance through increased outrage and heightened feelings of moral obligation. Contributing to this line of research, we will test how guilt and moral shame compare to other predictors of collective action and whether they are predictive of anti-government action in an authoritarian state over and above other emotions and attitude (RQ1).

@@

All study materials, including the questionnaire, data, and code, can be found on the Open Science Platform: https://osf.io/4pd2v/?view_only=23597fdc19424706818a72fc0df43009. The preregistration protocol is available at https://aspredicted.org/SGV_SBD.

@@

We collected a quota sample, representative of the Russian population by gender and age. Participants were recruited via the online crowdsourcing platform Yandex Toloka in August 2022.  This was the only feasible option to collect fully anonymized data in Russia, which was crucial, given that the repercussions for expressing opposition to the war can be severe. Participants received \$0.5 (30 RUB) for their participation. A total of N=`r length(d2$start)` participants gave full informed consent to participate in the study. Thirty participants were excluded since they dropped out of the study before completion, and another eight were excluded for failing more than one out of three attention checks. The effective sample size is N=`r length(d3$start)`.

@@

The sample was balanced by gender (`r round((gender_table[2]/length(d3$start)*100), digits = 1)`% women, `r round((gender_table[1]/length(d3$start)*100), digits = 1)`% men, `r round((gender_table[3]/length(d3$start)*100), digits = 1)`% non-binary or no response) and age cohort (`r age_range[1]` to `r age_range[2]` years old, M=`r round(mean(d3$age), digits = 0)`, SD=`r round(sd(d3$age), digits = 0)`). About 51% of the sample had a tertiary degree (BA or higher) and about 55% said they had lower than average income. All participants were Russian citizens and `r round((residence_table[1]/length(d3$start)*100), digits = 1)`% lived in Russia at the time of data collection. Most (`r round((ethn_table[1]/length(d3$start)*100), digits = 0)`%) identified as ethnic Russian, `r round((ethn_table[2]/length(d3$start)*100), digits = 0)`% as Tatar, about `r round((ethn_table[4]/length(d3$start)*100), digits = 0)`% as Ukrainian, Chuvash, and Bashkir each, and the remaining `r round((ethn_table[7]/length(d3$start)*100), digits = 0)`% chose the “other” category.

@@

***Anti-war actions***. We measured *past behavior* by asking how many of the eight political actions listed did the participant do to oppose the war in Ukraine (e.g., “signing a petition”, “participating in a demonstration”, etc.). The responses ranged from 0 to 8, with `r round(table(d3$pb_smop_opp)[1]/length(d3$pb_smop_opp)*100, digits = 0)`% of participants reporting zero actions. *Behavioral intention* to participate in any political action against the war was measured with 3 items (“I want to/intend to/plan to take part in political action against the war in Ukraine”), on a 5-point scale from 1 – ‘ Completely disagree’ to 5 – ‘Completely agree’ ($\alpha$ = `r round(alpha_bi_oppos$total[1],digits=2)`; `r round(table(means_bi)[1]/length(d3$bi_smop_opp1)*100, digits = 0)`% of participants had a mean score of 1). Finally, we asked about *the probability that the person will take any political action*, from 1 – highly improbable, 10 – (almost) certainly (`r round(table(d3$bp_smop_opp)[1]/length(d3$bp_smop_opp)*100, digits = 0)`% answered ‘1’).

***Attitude towards the war*** was measured with 3 items, on a bipolar 5-pt Osgood-style scale: “The special military operation in Ukraine is…”: bad-good, harmful-beneficial, useless-useful ($\alpha$ = `r round(alpha_SMO$total[1],digits=2)`, M = `r round(mean(d3$smop_att, na.rm = TRUE), digits = 2)`, SD = `r round(sd(d3$smop_att, na.rm = TRUE), digits = 2)`).

***Political beliefs.*** We used four different scales to capture participants’ beliefs about political agency and the role of democracy. All items were assessed on a 7-point scale from 1 – ‘Absolutely disagree’ to 7 – ‘Absolutely agree’.

*Support for democratic values* was measured with a 6-item scale from @CanettiNisim2007. Example item: "Every citizen has the right to take his convictions to the street if necessary" ($\alpha$ = `r round(alpha_demv$total[1],digits=2)`).

@@

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = `r round(summary5$fit[9], digits = 3)`, RMSEA = `r round(summary5$fit[17], digits = 3)`, SRMR = `r round(summary5$fit[25], digits = 3)`). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = `r round(summary6$fit[9], digits = 3)`, RMSEA = `r round(summary6$fit[17], digits = 3)`, SRMR = `r round(summary6$fit[25], digits = 3)`). This modified model includes the *political responsibility* factor with 5 items, *political alienation* factor with 3 items (the item “I believe public officials don't care much what people like me think” was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: “*politicians are evil*” (4 items, e.g., “Politicians are only interested in getting and maintaining power”) and “*politics is evil*” (4 items, e.g., “No man can hope to stay honest once he enters politics”).

***Positive and negative emotions.*** We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale [@Watson1988]. Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-pt scale (1 – ‘do not experience at all’ to 5 – ‘extremely’). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt and/or shame, respectively (we took this precaution because of negative prior experiences when using these measures in Russia; see @Grigoryan2018; @Grigoryan2017, for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame) using CFA and found the fit to be unsatisfactory (CFI = `r round(summary7$fit[9], digits = 3)`, RMSEA = `r round(summary7$fit[17], digits = 3)`, SRMR = `r round(summary7$fit[25], digits = 3)`). A 3-factor structure with positive emotions (joy, hope, enthusiasm), negative dominant emotions (anger, disgust, contempt) and negative submissive emotions (fear, sadness, depression) showed a satisfactory fit to the data (CFI = `r round(summary8$fit[9], digits = 3)`, RMSEA = `r round(summary8$fit[17], digits = 3)`, SRMR = `r round(summary8$fit[25], digits = 3)`) (see @Mehrabian1980, for the distinction between dominant and submissive emotions).

>>>>

---
title: "Guilt, shame, and anti-war action in an authoritarian country at war"
output:
  bookdown::pdf_document2:
    latex_engine: lualatex
    keep_tex: yes
    toc: no
  word_document:
    toc: no
  bookdown::word_document2:
    toc: no
  html_document:
    df_print: paged
always_allow_html: yes
test: "`r fontname <- 'Times New Roman'`
`r OS <- Sys.info()['sysname']`
`r if (OS=='Linux'){fontname <- 'Nimbus Roman'}`"
header-includes:
- \usepackage{indentfirst}
- \usepackage{sectsty}
- \allsectionsfont{\centering}
- \usepackage{fontspec}
- \usepackage{polyglossia}
- \setdefaultlanguage{english}
- \setotherlanguages{russian}
- \setmainfont{`r fontname`}
- \newfontfamily{\cyrillicfonttt}{`r fontname`}
- \usepackage{float}
- \usepackage{float} \floatplacement{figure}{H}
- \newcommand{\beginsupplement}{\setcounter{table}{0}
- \renewcommand{\thetable}{S\arabic{table}}
- \setcounter{figure}{0}
- \renewcommand{\thefigure}{S\arabic{figure}}}
- \usepackage{fancyhdr}
- \pagestyle{fancy}
indent: no
bibliography: references.bib
csl: elife.csl
---

\fancyhead[LH]{GUILT, SHAME, AND ANTI-WAR ACTION}
\fancyhead[RH]{}

@@

$^{5}$ Department of Psychological and Brain Sciences, University of Massachusetts Amherst
\vspace{40mm}

@@

\noindent This is a computationally reproducible version of the final paper. The version of record was published in *Political Psychology*, and is available at: https://doi.org/10.1111/pops.12985

\vspace{10mm}

@@

Feeling guilt and shame for the harm done to others by the ingroup can facilitate intergroup reconciliation. Most of the studies showing this effect are conducted in democratic countries and on historical, not current, conflicts. We investigated the role of group-based guilt and shame in collective action in an authoritarian country at war. We asked more than 1000 Russians living in Russia, in a sample representative of the country’s population by gender and age, about their experiences of group-based guilt and shame regarding Russia’s invasion of Ukraine and their past and future anti-war political actions. We tested whether political efficacy is necessary for experiencing group-based guilt and shame, and whether these emotions are predictive of anti-war action over and above other emotions and attitudes. Democratic values, not political efficacy, were the most robust predictor of group-based guilt and shame. Only moral shame, but not image shame or guilt predicted past and future anti-war action. Whereas attitude towards the war and moral shame predicted engagement in anti-war action (vs none), other negative dominant emotions like anger predicted the degree of this engagement. We highlight the gaps in the study of collective action and the need for more evidence from non-democratic contexts.

\vspace{5mm}
*Keywords*: group-based emotions, guilt, moral shame, image shame, collective action, Russia-Ukraine war, authoritarianism

@@

packagelist <- c('readr','psych','GPArotation','lavaan','pscl','ggplot2','sjPlot','apaTables', 'car','dplyr','MASS','writexl','readxl','kableExtra','rstatix','tinytex','GGally','ggthemes')
missingpackages <- packagelist[!packagelist %in% installed.packages()[,1]]
if (length(missingpackages)>0){install.packages(missingpackages)}
toinstall <- packagelist[which(!packagelist %in% (.packages()))]
invisible(lapply(toinstall,library,character.only=TRUE))

@@

pdf('corplot.pdf', width=8, height=8)
plot <- d3[,c(216,218,220,170,251,182)]
colnames(plot) <- c('Guilt','Moral shame','Image shame', 'Past behavior',
                    'Behavioral intention', 'Action probability')

ggpairs(plot, lower=list(continuous=wrap("smooth",alpha=0.5,size=1,color="darkblue")),
        diag=list(continuous = wrap("densityDiag",size=1,color="darkblue")), 
        upper=list(continuous="cor", axisLabels='show'))
dev.off()

@@

On February 24, 2022, Russia, the country with the biggest nuclear warhead inventory in the world, invaded the neighboring Ukraine. Although the Russia-Ukraine war started in 2014 with the occupation of Crimea, Russian officials consistently denied their involvement in the region. Eight years later, Russia openly invaded Ukraine on four fronts: North (Kyiv), North-East (Kharkiv), South-East (Donetsk and Luhansk), and South (Crimea). The official ideological narrative supporting the invasion was provided in a televised speech by Vladimir Putin preceding the invasion by mere minutes. In this speech, Putin falsely claimed that Ukraine is governed by a group of neo-Nazis who hold the population hostage and persecute the Russian speakers in the country. The announced goals of this “special military operation” (the use of the word “war” soon after will be criminalized in Russia) were the “denazification” and “demilitarization” of Ukraine.

In the first days following the invasion, thousands of Russians took to the streets to protest. However, nearly 15.000 people were detained in the first weeks following the invasion [@OVDInfo2022] and the street protests quickly died down. New laws were introduced regularly to further suppress protest, among them the law about the “discreditation of the Russian armed forces”, which essentially prohibited any anti-war statements or actions [@Eckel2022]. After the introduction of this law, the few remaining independent media outlets either stopped covering the war or left the country [@MoscowTimes2022].

Against the backdrop of the reports of war crimes committed by the Russian armed forces in Ukraine and the repression of all dissent in Russia, the discussions among the well-educated Russians, many of whom left the country, often revolved around the topics of collective responsibility, guilt, and shame [e.g., @Grigoryan2022; @Treus2022]. As these discussions took shape, it became clear that the majority view, even among the Russian “intelligentsia” that opposed the war, is that feelings of guilt and shame are inappropriate or even harmful [@Kynev2022]. For example, in a series of interviews with prominent Russian intellectuals and artists, Ekaterina Gordeeva, a journalist with more than 1.5 million subscribers on YouTube, asked each of her guests whether they feel guilt or shame for what is happening in the country; with very few exceptions, the answer was “no” [@Gordeeva2022]. The push-back against the idea of feeling guilt or shame for what your country is doing was so strong that soon a new hashtag #IamNotAshamed (#МнеНеСтыдно) became popular across social media platforms.

Research on group-based guilt and shame and collective action has been predominantly carried out in Western democracies [with notable exception of a number of studies on the Israeli-Palestinian conflict; @Roccas2006; @WeissKlayman2020]. Although there is strong evidence that experiencing these emotions can motivate people to engage in collective action and support intergroup reconciliation efforts and oppose military actions by one's country [@Hakim2021; @Iyer2007; @Lickel2011], research on preconditions for experiencing these emotions and on the power of these emotions to motivate collective action in authoritarian regimes is lacking. One could argue that when people have no control, or believe that they have no control, over the decisions of their government, feelings of guilt and shame would not arise or would not be associated with behavioral outcomes, as they are in democratic contexts, where these emotions serve the function of regulating ingroup’s moral behavior. If the ingroup’s behavior cannot be changed, then these emotions might lose their functionality and be downregulated. Is the sense that one has some control over the political events in their country a necessary precondition for feeling group-based guilt and shame? Are these feelings relevant in authoritarian regimes? Can they motivate behavior in such contexts? We set out to address these questions in Russia, an authoritarian country at war.

@@

This prediction that particular moral emotions are linked uniquely to different forms of intergroup behavior (with guilt linked to reconciliatory behavior and shame linked to avoidance) is not well supported by the empirical evidence available so far. A recent meta-analysis of group-based emotions ([@Hakim2021], 101 effect sizes, 58 samples, N = 10,305) found that guilt, shame, and anger are equally predictive of support for reparations, with a strong average effect size of r = .44 and no significant differences between emotions. Furthermore, theHow can this inconsistency regarding the impact of shame on reconciliatory behaviors be explained?

The distinction between moral shame and image shame [@Allpress2010; @Rees2013] helps explain which elements of shame are predictive of reconciliatory behaviors and which are predictive of avoidant behaviors. This distinction is made based on different appraisals that the feeling of shame can be accompanied with: we feel moral shame when the group members’ behavior violates important moral standards or values of the group, and we feel image shame when the group members’ behavior tarnishes the ingroup’s reputation [@Rees2013]. Studies that make the distinction between moral shame and image shame find that moral shame is predictive of reconciliatory behaviors on par or more strongly than guilt, whereas image shame is predictive of distancing and avoidance [@Allpress2010; @Allpress2014; @Grigoryan2017; @Nooitgedagt2021; @Rees2013]. We therefore expect that moral shame and guilt, but not image shame, will predict a stronger intention to act against the war in Russia (H1).

## Acknowledgement of responsibility as a precondition for experiencing guilt and shame {.unnumbered}

Although rarely studied, the acknowledgement of collective responsibility for the ingroup’s wrongdoings could be a necessary precondition for experiencing group-based guilt and shame. In a series of experiments, @CehajicClancy2011 showed that guilt mediates the effect of an intervention designed to increase acknowledgement of responsibility for the ingroup’s wrongdoing and support for reparations. Consistent with this finding, a recent set of studies showed that beliefs of group malleability shape experiences of guilt and shame [@WeissKlayman2020]: participants who believed that groups can change and those who thought that others believe that groups can change (meta-beliefs) showed higher levels of guilt and shame.

@@

Various definitions and operationalizations have been suggested for both political cynicism and political alienation constructs, but broadly, both capture a belief that the political process is run without input from ordinary citizens, that politics and politicians can’t be trusted, and that political participation is futile [@Litt1963; @Schwartz1973]. Both political alienation and cynicism have been shown to predict political non-participation [@Adams2006; @Pinkleton2004]. As political alienation and cynicism would indicate disengagement and dis-identification with the political process, we predict that higher levels of alienation and cynicism would predict lower levels of group-based shame and guilt. It is possible that political cynicism could be associated with anti-government attitudes and behavior, but only when operationalized as mistrust of the current government [@Erber1990]. When operationalized as apathy and withdrawal from the political process, we expect the effects of political alienation and cynicism in authoritarian contexts would be similar to those in democratic contexts.

Conversely, democratic values and the notion of political responsibility highlight the importance of political participation. Democratic values represent support for key democratic principles such as freedom of speech, minority rights, and the rule of law [@CanettiNisim2007], and political responsibility represents a belief that people have a responsibility to participate in the political process. Democratic values have been repeatedly shown to predict political participation [@Chang2017; @Karp2015; @Oser2018]. We predict that higher levels of support for democratic values and political responsibility would predict higher levels of group-based shame and guilt because both support for democratic values and political responsibility beliefs imply engagement with the political process. One could argue that the conflict between democratic values and a non-democratic political process in an authoritarian state would lead to dis-identification and political disengagement among those who strongly support democratic values. However, we believe it more likely that the disengagement of people from politics in authoritarian states is achieved (besides repressions) by a combination of making democracy seem less desirable and of corrupting the meaning of what democracy means [see the distinction between liberal and authoritarian notions of democracy, @Kirsch2018]. We would expect that in any authoritarian state, those most motivated to change the status quo, and therefore those who would feel most negative emotions about the wrongdoings of their state, would be those who support liberal democratic values.

@@

The integrative social identity model of collective action (SIMCA) – perhaps the most comprehensive and widely used model of collective action in psychology – identifies three key components of collective action: perceived injustice, identity, and efficacy [@Zomeren2008]. In a meta-analysis of more than 180 independent samples, @Zomeren2008 found that each of these three components uniquely predict collective action, with similar effect sizes. As the diversity of samples and issues studied grew in recent years, authors extended the model by introducing moral beliefs as a key component of the model [@Zomeren2018]. The updated meta-analysis [@Agostini2021] introduces the dual chamber model of collective action, which places identity and morality as two interrelated and equally fundamental motivational drivers of collective action that translate into action via perceptions or feelings of injustice and efficacy.

This meta-analysis [@Agostini2021] incorporates 1.235 effects from 403 samples with a total sample size of over 120.000. Among other things, the meta-analysis looks into structural constraints on motivations for collective action. Although many limitations remain regarding diversity of samples and issues studied, and political regime or the democracy index are not included in the analysis, some variables can serve as indirect indicators of how restrictive the social context is. These include Western vs. non-Western samples (assuming the West is on average more democratic), collectivism cultural dimension from Hofstede (assuming behavior is regulated more strongly in collectivist societies), and hierarchy – egalitarianism cultural dimension from Schwartz (assuming more hierarchical societies are more restrictive). Keeping in mind all the limitations of such approximation, it is worth noting that the effect of morality in collective action was stable across all moderator variables, whereas the effect of identity varied considerably and was weaker in non-Western, collectivistic, and more hierarchical societies. This suggests that the content of identity, including who is defined as “us”, what it means to be “us”, and how politicized this identity is can vary considerably across contexts. Based on these meta-analytical findings, we would expect morality, i.e., moral shame, to be the most robust predictor of collective action in the Russian context.

This expectation is supported by other studies that looked specifically at collective action in non-democratic contexts. In a series of studies in five non-democratic states (Egypt and Turkey in 2013, Russia, Ukraine, and Hong Kong in 2014), Ayanian and colleagues [@Ayanian2016; @Ayanian2021] show how perceptions of risk in repressive contexts can have counter-intuitive effects on willingness to participate in collective action. While perceived risk can quell collective action through increased fear, it can also spur resistance through increased outrage and heightened feelings of moral obligation. Contributing to this line of research, we will test how guilt and moral shame compare to other predictors of collective action and whether they are predictive of anti-government action in an authoritarian state over and above other emotions and attitudes (RQ1).

@@

The study was approved by the ethics committee of the Psychology Department at [Anonymized university]. All study materials, including the questionnaire, data, and code, can be found on the Open Science Framework platform: https://osf.io/4pd2v/?view_only=23597fdc19424706818a72fc0df43009. The preregistration protocol is available at https://aspredicted.org/SGV_SBD. The scope of the preregistration protocol is broader than the scope of the current paper: it includes predictions regarding group-based guilt and shame, positive and negative emotions more broadly, and social identity as potential predictors of political behavior. In this manuscript, we focus on predictions regarding group-based guilt and shame and report the analyses pertaining to positive and negative emotions in the supplement. The hypotheses regarding the role of social identity (global-local identity and constructive and blind patriotism) will be the focus of a future publication.

@@

We collected a quota sample, representative of the Russian population by gender and age. Participants were recruited via the online crowdsourcing platform Yandex Toloka in August 2022 and completed the survey anonymously.  This was the only feasible option to collect fully anonymized data in Russia, which was crucial, given that the repercussions for expressing opposition to the war can be severe. Participants received \$0.5 (30 RUB) for their participation. A total of N=`r length(d2$start)` participants gave full informed consent to participate in the study. Thirty participants were excluded since they dropped out of the study before completion, and another eight were excluded for failing more than one out of three attention checks. The effective sample size is N=`r length(d3$start)`.

@@

The sample was balanced by gender (`r round((gender_table[2]/length(d3$start)*100), digits = 1)`% women, `r round((gender_table[1]/length(d3$start)*100), digits = 1)`% men, `r round((gender_table[3]/length(d3$start)*100), digits = 1)`% non-binary or no response) and age cohort (`r age_range[1]` to `r age_range[2]` years old, M=`r round(mean(d3$age), digits = 0)`, SD=`r round(sd(d3$age), digits = 0)`). About 55% said they had lower than average income and about 51% of the sample had a tertiary degree (BA or higher). For comparison, 63% of 25-34 year-old Russians have tertiary education according to the latest available OECD data from 2018 (OECD, 2019). All participants were Russian citizens and `r round((residence_table[1]/length(d3$start)*100), digits = 1)`% lived in Russia^[We additionally tested the robustness of our findings when excluding the participants who did not reside in Russia at the time of data collection. None of the results were affected by this exclusion.] at the time of data collection. Most (`r round((ethn_table[1]/length(d3$start)*100), digits = 0)`%) identified as ethnic Russian, `r round((ethn_table[2]/length(d3$start)*100), digits = 0)`% as Tatar, about `r round((ethn_table[4]/length(d3$start)*100), digits = 0)`% as Ukrainian, Chuvash, and Bashkir each, and the remaining `r round((ethn_table[7]/length(d3$start)*100), digits = 0)`% chose the “other” category.

@@

The means, standard deviations, and correlations between all variables are available in Table S1 of the online supplement. Figure 1 show the distribution and correlations between group-based emotions and past and future anti-war action.

***Political beliefs.*** We used four different scales to capture participants’ beliefs about political agency and the role of democracy. All items were assessed on a 7-point scale from 1 – (*absolutely disagree*) to 7 – (*absolutely agree*).

*Support for democratic values* was measured with a 6-item scale from @CanettiNisim2007. Example item: "Every citizen has the right to take his convictions to the street if necessary" ($\alpha$ = `r round(alpha_demv$total[1],digits=2)`; M=`r `, SD=`r `).

@@

Since the constructs of political alienation, cynicism, and responsibility are closely related and none of these scales were used in a Russian sample before, we conducted a confirmatory factor analysis (CFA) to test whether the 3-factor structure could be confirmed. The 3-factor model showed unsatisfactory fit to the data (CFI = `r round(summary5$fit[9], digits = 3)`, RMSEA = `r round(summary5$fit[17], digits = 3)`, SRMR = `r round(summary5$fit[25], digits = 3)`). An exploratory factor analysis suggested that a 4-factor model, with the political cynicism scale split in two, fits the data better. CFA confirmed this result: a model with 4 factors showed a good fit to the data (CFI = `r round(summary6$fit[9], digits = 3)`, RMSEA = `r round(summary6$fit[17], digits = 3)`, SRMR = `r round(summary6$fit[25], digits = 3)`). This modified model includes the *political responsibility* factor with 5 items (M=`r`, SD=`r`), *political alienation* factor with 3 items (M=`r`, SD=`r`; the item “I believe public officials don't care much what people like me think” was removed since it loaded on both political alienation and cynicism subscales), and two facets of political cynicism: “*politicians are bad*” (4 items, e.g., “Politicians are only interested in getting and maintaining power”; M=`r`, SD=`r`) and “*politics is bad*” (4 items, e.g., “No man can hope to stay honest once he enters politics”; M=`r`, SD=`r`).

***Positive and negative emotions.*** We measured emotions that participants experience when thinking about the war using the short version of the Positive and Negative Affect Scale [PANAS; @Watson1988]. Twelve emotions (fear, joy, sadness, hope, anger, enthusiasm, disgust, pride, contempt, depression, guilt, and shame) were rated on a 5-point scale from 1 (*do not experience at all*) to 5 (*extremely*). If participants answered 2 or higher to the guilt and shame questions, they were shown questionnaires for group-based guilt (n=) and/or shame (n=), respectively. We took this precaution participants in earlier studies conducted in Russia complained about the double-barreled nature of items measuring group-based guilt and shame, where they felt that the questions imply that participants experience some level of guilt or shame (see @Grigoryan2018; @Grigoryan2017, for details). We tested the two-factor structure of the PANAS scale (positive and negative affect, excluding guilt and shame^[We excluded guilt and shame since we planned to estimate whether group-based guilt and shame are predictive of behavior over and above other emotions. Including PANAS measures of guilt and shame into the index of negative emotions would have confounded the two measures and prevented us from addressing this question (RQ1).]) using CFA and found the fit to be unsatisfactory (CFI = `r round(summary7$fit[9], digits = 3)`, RMSEA = `r round(summary7$fit[17], digits = 3)`, SRMR = `r round(summary7$fit[25], digits = 3)`). A 3-factor structure with *positive emotions* (joy, hope, enthusiasm; M=, SD=), *negative dominant emotions* (anger, disgust, contempt; M=, SD=) and *negative submissive emotions* (fear, sadness, depression; M=, SD=) showed a satisfactory fit to the data (CFI = `r round(summary8$fit[9], digits = 3)`, RMSEA = `r round(summary8$fit[17], digits = 3)`, SRMR = `r round(summary8$fit[25], digits = 3)`; see @Mehrabian1980 for the distinction between dominant and submissive emotions).

***Group-based guilt and shame.*** We used the measures of group-based guilt, image shame, and moral shame developed by @Rees2013, Russian adaptation by @Grigoryan2017. All items were answered on a 5-point scale from 1 (*completely disagree*) to 5 (*completely agree*). Guilt was measured with 3 items (e.g., “I feel guilty for the manner in which Ukrainian people have been treated by Russians”, $\alpha$ = `r round(alpha_guilt$total[1],digits=2)`), image shame with 5 items (“I feel ashamed when I realize that other countries might think of Russia negatively because of our involvement in Ukraine”, $\alpha$ = `r round(alpha_ishame$total[1],digits=2)`), and moral shame with 4 items (“I feel ashamed because Russia’s actions with regard to Ukraine have been immoral”, $\alpha$ = `r round(alpha_mshame$total[1],digits=2)`). All missing values were replaced with zeros, since participants who didn’t see the scale were the ones who reported no feelings of guilt or shame in the PANAS. With this imputation, we were able to use the whole dataset without excluding those who reported experiencing no guilt or shame. Mean values after this imputation were M=1.23 (SD=1.69) for guilt, M=1.13 (SD=1.70) for image shame, and M=1.09 (SD=1.67) for moral shame.
